### 2016-05-24 thesis/methods/methods.tex
tmp commit

Samples were classified based on the \gls{who} definition, as shown in Table~\ref{tab:whobmiclass}.
\begin{table}[hb]
	\caption{\gls{who} defined \gls{bmi} classification}
	\label{tab:whobmiclass}
	\begin{center}
		\begin{tabular}{lc}
			\bfseries {Classification} & \bfseries {\gls{bmi} Value}\\
			\hline
			\rule{0pt}{2.25ex}Underweight & \textless{} 20.0\\
			Normal weight/lean & 20.0$\sim$24.9\\
			Overweight & 25.0$\sim$29.9\\
			Obese & \textgreater{} 30.0\\
		\end{tabular}
	\end{center}
\end{table}

@@

The clinical data for all cancer types (33 types in total) were downloaded from \gls{tcga} database and were checked for the height and weight data for each sample, for each cancer type.
Any cancer type with no weight and/or height data of the samples were excluded from the project, as no \gls{bmi} information can be obtained without these data.

This gave a total of eight cancer types (\gls{blca}, \gls{cesc}, \gls{coad}, \gls{kirp}, \gls{lihc}, \gls{read}, \gls{skcm}, \gls{ucec}) to be used for further analyses.

@@

RNA-seq data for the eight cancer types with sample \gls{bmi} information available were downloaded from the \gls{icgc} database.
The RNA-seq data for each of the eight cancer types were imported into R.

\section{Validation of methodology used by \citet{Creighton2012}}
\label{sec:methodvalidation}

@@

The raw microarray gene expression data files of all the samples used by \citet{Creighton2012} were downloaded from the \gls{geo} database.
Raw data were imported into R/RStudio for further analyses.

(add data/sample composition -- ethinicity, grade, etc -- maybe in table form)

@@

The raw microarray data for all 103 samples in the study were normalised using \gls{rma} method from the \textit{affy} package in R.
The data was then standardised so that each of the genes had a mean of 0 and a standard deviation of 1.
(mention scaling anything over 3 were made to be 3 for heatmap scaling)

@@

799 obesity associated gene probes identified by \citet{Creighton2012} were downloaded from (where? - i think the same place as microarray data).
The list of gene probes were imported into R for further analyses and validation of methodology.

\subsection{Differential gene expression analysis}
\label{subsec:deg}

All 103 samples from the \citet{Creighton2012} study were split into two groups (obese and non-obese) based on the sample \gls{bmi} status as defined by the \gls{who} classification (see Table~\ref{tab:whobmiclass}).
The difference in the gene expression between these two sample groups were analysed by fitting a linear model to the data, using the \textit{limma} package in R.
The top 799 differentially expressed gene probes that had p-value \textless{} 0.01 were identified.

@@

The obesity associated gene probes identified by \citet{Creighton2012} were compared with each of the three sets of gene probes identified from the differential gene expression analysis (see \ref{subsec:deg} \nameref{subsec:deg}).
The presence of any of the \citet{Creighton2012} genes were checked for in each of the gene sets identified at a p-value \textless{} 0.05, as well as the top 799 gene probes at p-Value \textless{} 0.01.

>>>>

Samples were classified based on the \gls{who} definition, as shown in Table~\cref{tab:whobmiclass}.
\begin{table}[hb]
	\caption{\gls{who} defined \gls{bmi} classification}
	\label{tab:whobmiclass}
	\begin{center}
		\begin{tabular}{lc}
			\bfseries {Classification} & \bfseries {\gls{bmi} Value}\\
			\hline
			\rule{0pt}{2.25ex}Underweight & \textless{} 20.0\\
			Normal weight/lean & 20.0$\sim$24.9\\
			Overweight & 25.0$\sim$29.9\\
			Obese & \textgreater{} 30.0\\
		\end{tabular}
	\end{center}
\end{table}

@@

The clinical data for all available cancer types (33 types in total) were downloaded from \gls{tcga} database and were checked for the height and weight data for each sample, for each cancer type.
Any cancer type with no weight and/or height data of the samples were excluded from the project, as no \gls{bmi} information can be obtained without these data.

This gave a total of eight cancer types: \gls{blca}, \gls{cesc}, \gls{coad}, \gls{kirp}, \gls{lihc}, \gls{read}, \gls{skcm}, \gls{ucec}.
All eight cancer types were used for further analyses.

@@

\gls{rnaseq} data for the eight cancer types with sample \gls{bmi} information available were downloaded from the \gls{icgc} database.
The \gls{rnaseq} data for each of the eight cancer types were imported into \gls{r}.

%TODO: change the title name?
\section{\gls{icgc} sample exclusion criteria}
\label{sec:sampleexclusion}

Samples from the downloaded data were removed using the following criteria.

\subsection{Sample BMI availability}
\label{subsec:samplebmiavailability}

Some of the samples did not have either height or weight data, or both.
These samples were removed from the analyses.

\subsection{Sample ID conversion and matching}
\label{subsec:sampleidconversion}

%TODO: check the actual ID stripping process
%TODO: check the WGCNA function and check if it uses median/mean/whatever for choosing the sample

\gls{icgc} sample ID in the \gls{rnaseq} data were stripped so that the ID from the \gls{rnaseq} data matched with the \gls{tcga} ID in the clinical data.
After stripping the ID, samples were checked for any overlaps.
Where there was an overlap in the sample ID in the \gls{rnaseq} data, a single sample was chosen to represent that particular ID by using the \texttt{collapseRows} function in the \textit{WGCNA} package in \gls{r}.

In each of the eight cancer types, the \gls{tcga} sample ID in the clinical data were cross-checked with the \gls{tcga} sample ID in the \gls{rnaseq} data, and vice versa.
Any sample that did not have either clinical or \gls{rnaseq} data were removed, so both clinical and \gls{rnaseq} data were available for all of the samples that were included in the analyses.
See \cref{tab:samplesize} for a summary of the number of samples included in the analyses for each cancer type.

\begin{table}[h]
	\caption{Summary of the total number of samples included in the analyses for each cancer type}
	\label{tab:samplesize}
	\begin{center}
		\begin{tabular}{cc}
			\textbf{Cancer Type}   & \textbf{Total} \\
			\hline
			\rule{0pt}{2.25ex}BLCA & 261   \\
			CESC                   & 224   \\
			COAD                   & 226   \\
			KIRP                   & 124   \\
			LIHC                   & 264   \\
			READ                   & 73    \\
			SKCM                   & 218   \\
			UCEC                   & 482   \\
		\end{tabular}
	\end{center}
\end{table}

\section{Data processing -- \citet{Creighton2012} data}
\label{sec:dataprocessingcr}

@@

The raw microarray gene expression data and clinical data files of all *** samples from the \citet{Creighton2012} study were downloaded from the \gls{geo} database (include website name/GEO ID here).
Raw data were imported into \gls{r}/RStudio for subsequent analyses.

%TODO: add data/sample composition -- ethinicity, grade, etc -- maybe in table form

@@

The raw microarray data for all 103 samples from the study were normalised using \gls{rma} method from the \textit{affy} package in \gls{r}.
The data was then standardised so that each of the genes had a mean ($\mu$) of 0 and a standard deviation ($\sigma$) of 1.
This ensured that all of the genes were on the same scale as one another, allowing for comparison.

@@

799 obesity associated gene probes identified by \citet{Creighton2012} were downloaded from (supplementary data??).
The list of gene probes were imported into \gls{r}.
The gene probe IDs were converted into gene symbols using the \textit{hgu133a.db} package in \gls{r}.
Any duplicates and any gene symbols that were not present in both the Creighton data and the \gls{icgc} data were removed, which left *** unique obesity associated genes.

\section{Validation of results -- \citet{Creighton2012}}
\label{sec:valresultscr}

The results presented by \citet{Creighton2012} were validated using the same raw data they have used in their study.
See  \cref{sec:dataprocessingcr} for details on how the data was processed before the analyses.

\subsection{Metagene and sample gene expression pattern}
\label{subsec:crmetage}

The gene probe IDs were converted into gene symbols using the \textit{hgu133a.db} package in \gls{r}.
Where there was a duplicated gene in the data, appropriate gene data was chosen to represent that particular gene by using the \texttt{collapseRows} function in the \textit{WGCNA} package in \gls{r}.
All of the *** obesity associated genes from  \cref{subsec:crobsgene} were selected from the data.

\gls{svd} was applied to this data to obtain the decomposed matrices of the data.
%TODO: how do I explain how I got the \gls{metagene} scores?
The \gls{metagene} scores of the samples were obtained by taking the first principal component of the (eigenvector...?).
The \gls{metagene} scores were ranked based on the number of samples in the data (103 samples in Creighton data), and the values of the scores were between 0 and 1.

These \gls{metagene} scores were plotted together with the gene expression of the obesity associated gene probes for each sample on a heatmap.
The samples in the heatmap were ordered based on their ranked \gls{metagene} scores.
To have a better colour scales in the heatmap, any expression values that were greater than 3 or less than -3 were assigned the value of 3 and -3, respectively.

\subsection{Metagene and sample \gls{bmi}}
\label{subsec:crmetabmi}

The same \gls{metagene} scores as in the previous section (\cref{subsec:crmetage}) were used to compare the \gls{metagene} scores with the samples' \gls{bmi} statuses and values.
Boxplot was plotted to show the association of the \gls{metagene} scores with the sample \gls{bmi} status, and scatter plot was plotted to show the association with the sample \gls{bmi} value.

%TODO: check the formal definition of these test/values
For the boxplot, the p-value was obtained from the ANOVA test.
For the scatter plot, the regression line was plotted and the adjusted $R^2$ value was calculated for the regression line.

\section{Creighton metagene in \gls{icgc} cancer data}
\label{sec:crmetaicgc}

\subsection{\gls{icgc} data processing}
\label{subsec:datprocicgc}

The obesity associated genes from \cref{subsec:crobsgene} were selected in each of the eight cancer types.
All cancer data had 1 added to the count data, so no infinite values were introduced when the data was logged.
All cancer data were logged to the base 10.

All of the data were then standardised so that each of the genes had a mean ($\mu$) of 0 and a standard deviation ($\sigma$) of 1.
Any ``NaN'' values in the data introduced from this were assigned a value of 0.

\subsection{Transformation matrix}
\label{subsec:transmatcr}

Transformation matrix was formed from the decomposed matrices that resulted from the application of \gls{svd} to the Creighton data (see \cref{subsec:crmetage}).
%TODO: add the derivation of transformation matrix??

\subsection{Transformed \gls{icgc} data and metagene}
\label{subsec:icgctranscr}

Each of the eight cancer data from \cref{subsec:datprocicgc} were transformed using the transformation matrix from \cref{subsec:transmatcr}.
Metagene scores of the samples in each cancer types were obtained by taking the first column of the cancer matrix that resulted from the transformation.
The \gls{metagene} scores were ranked based on the number of samples in the data, and the scores ranged from 0 to 1.

\subsection{Creighton metagene and sample gene expression in \gls{icgc} data}
\label{subsec:crmetaicgcge}

\Gls{metagene} scores from \cref{subsec:icgctranscr} were plotted together with the gene expression of the obesity associated gene probes for each sample on a heatmap, for each cancer type.
The samples in the heatmap were ordered based on their ranked \gls{metagene} scores.
To have a better colour scales in the heatmap, any expression values that were greater than 3 or less than -3 were assigned the value of 3 and -3, respectively.

\subsection{Creighton metagene and sample \gls{bmi} in \gls{icgc} data}
\label{subsec:crmetaicgcge}

The same \gls{metagene} scores as in the previous section were used to compare the \gls{metagene} scores with the samples' \gls{bmi} statuses and values for each cancer type.
Boxplot was plotted to show the association of the \gls{metagene} scores with the sample \gls{bmi} status, and scatter plot was plotted to show the association with the sample \gls{bmi} value.

%TODO: check the formal definition of these test/values
For the boxplot, the p-value was obtained from the ANOVA test.
For the scatter plot, the regression line was plotted and the adjusted $R^2$ value was calculated for the regression line.

%TODO: Not sure if I should cut this section to "see previous section for details"
\section{Data processing -- \citet{Fuentes-Mattei2014} data}
\label{sec:dataprocessingfm}

\subsection{Raw data from the \citet{Fuentes-Mattei2014} study}
\label{subsec:rawdatafm}

The raw microarray gene expression data files of all *** samples from the  \citet{Fuentes-Mattei2014} study were downloaded from the \gls{geo} database (include website name/GEO ID here).
Raw data were imported into \gls{r}/RStudio for subsequent analyses.

%TODO: add data/sample composition -- ethinicity, grade, etc -- maybe in table form

\subsection{Normalisation of raw \citet{Fuentes-Mattei2014} data}
\label{subsec:normfmdata}

The raw microarray data for all *** samples from the study were normalised using \gls{rma} method from the \textit{affy} package in \gls{r}.
The data was then standardised so that each of the genes had a mean ($\mu$) of 0 and a standard deviation ($\sigma$) of 1.
This ensured that all of the genes were on the same scale as one another, allowing for comparison.

\subsection{Obesity associated gene probes from the \citet{Fuentes-Mattei2014} study}
\label{subsec:fmobsgene}

130 obesity associated gene probes identified by \citet{Fuentes-Mattei2014} were downloaded from (supplementary data??).
The list of gene probes were imported into \gls{r}.
Any gene symbols that were not present in the Fuentes-Mattei data, the Creighton data and the \gls{icgc} data were removed, which left *** unique obesity associated genes.

\section{Validation of results -- \citet{Fuentes-Mattei2014}}
\label{sec:valresultsfm}

The results presented by \citet{Fuentes-Mattei2014} were validated using the \citet{Creighton2012} data, as the clinical data of the samples were not available.
See \cref{subsec:rawdatacr,subsec:normcrdata} for details on how the data was processed before the analyses.

\subsection{Fuentes-Mattei transformation matrix}
\label{subsec:transmatfm}

\gls{svd} was applied to Fuentes-Mattei data to obtain the decomposed matrices of the data, and this was used to make the transformation matrix.
%TODO: add the derivation of transformation matrix??

\subsection{Fuentes-Mattei metagene and gene expression pattern in Creighton data}
\label{subsec:fmmetage}

The gene probe IDs in Creighton data were converted into gene symbols using the \textit{hgu133a.db} package in \gls{r}.
Where there was a duplicated gene, appropriate gene data was chosen to represent that particular gene by using the \texttt{collapseRows} function in the \textit{WGCNA} package in \gls{r}.
All of the *** obesity associated genes from  \cref{subsec:fmobsgene} were selected from the data.

%TODO: how do I explain how I got the \gls{metagene} scores?
The \gls{metagene} scores of the samples were obtained by taking the first principal component of the transformed matrix.
The \gls{metagene} scores were ranked based on the number of samples in the data (*** samples in \citet{Creighton2012} data), and the scores ranged from 0 to 1.

These \gls{metagene} scores were plotted together with the gene expression of the obesity associated gene probes for each sample on a heatmap.
The samples in the heatmap were ordered based on their ranked \gls{metagene} scores.
To have a better colour scales in the heatmap, any expression values that were greater than 3 or less than -3 were assigned the value of 3 and -3, respectively.

\subsection{Metagene and sample \gls{bmi}}
\label{subsec:fmmetabmi}

The same \gls{metagene} scores from the previous section were used to compare the \gls{metagene} scores with the samples' \gls{bmi} statuses and values.
Boxplot was plotted to show the association of the \gls{metagene} scores with the sample \gls{bmi} status, and scatter plot was plotted to show the association with the sample \gls{bmi} value.

%TODO: check the formal definition of these test/values
For the boxplot, the p-value was obtained from the ANOVA test.
For the scatter plot, the regression line was plotted and the adjusted $R^2$ value was calculated for the regression line.

\section{Fuentes-Mattei metagene in \gls{icgc} cancer data}
\label{sec:fmmetaicgc}

\section{Differential gene expression analysis}
\label{sec:deg}

All 103 samples from the \citet{Creighton2012} study were split into two groups (obese and non-obese) based on the sample \gls{bmi} status as defined by the \gls{who} classification (see \cref{tab:whobmiclass}).
The difference in the gene expression between these two sample groups were analysed by fitting a linear model to the data, using the \textit{limma} package in \gls{r}.
The top 799 differentially expressed gene probes that had p-value \textless{} 0.01 were identified.

@@

The obesity associated gene probes identified by \citet{Creighton2012} were compared with each of the three sets of gene probes identified from the differential gene expression analysis (see \cref{sec:deg} \nameref{sec:deg}).
The presence of any of the \citet{Creighton2012} genes were checked for in each of the gene sets identified at a p-value \textless{} 0.05, as well as the top 799 gene probes at p-value \textless{} 0.01.
